Follow the Flow of Rain and Paint: An Instance Optimal Contour Tree Algorithm

### 2014-06-11 paint-contour-tree.tex
done with first very preliminary pass

\newcommand{\col}{col}
\newcommand{\mcol}{mcol}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\lift}{{\tt lift}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{R}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

\section{Preliminaries}
Consider a piecewise-linear (PL) $2$-manifold $\MM$ with boundaries.
This is represented by a triangulation (given as a DCEL) where edges
can be oriented and the boundary of a face is always counterclockwise. Some faces will \emph{not}
be triangular, and each of these corresponds to a \emph{boundary face}.
Let $G(\MM) = (V,E)$ denote this triangulation. 
We will assume that $\MM$ is embedding in $\RR^d$.
The manifold within each face is just linear.
We define a \emph{height function} on $\MM$, $f:\MM \mapsto \RR$. For each point in $x \in \MM$, let $f(x)$ simply be the value of $x_1$.
We distinguish between vertices and points in $\MM$. A point simply denotes any $x \in \MM$. A vertex is a point
corresponding to some vertex of the triangulation $G(\MM)$.

@@

@@

\smallskip
\begin{compactenum}
	\item Let $h$ be head of $\st$.
	\item While $\col(h) \setminus \mcol(h)$ non-empty:
	\begin{compactenum}
		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
		\item Obtain highest point $x$ in $T(c')$. 
		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
		and set $\mcol(x) = \{c'\}$.
		\item Let $h$ be new head of $\st$.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

@@

>>>>

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\h}{high}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{R}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

\section{Preliminaries}
Consider a piecewise-linear (PL) $2$-manifold $\MM$ with boundaries.
This is represented by a triangulation (given as a DCEL) where edges
can be oriented and the boundary of a face is always counterclockwise. Some faces will \emph{not}
be triangular, and each of these corresponds to a \emph{boundary face}.
Let $G(\MM) = (V,E)$ denote this triangulation. We assume this triangulation
has bounded degree.
We will assume that $\MM$ is embedding in $\RR^d$.
The manifold within each face is just linear.
We define a \emph{height function} on $\MM$, $f:\MM \mapsto \RR$. For each point in $x \in \MM$, let $f(x)$ simply be the value of $x_1$.
We distinguish between vertices and points in $\MM$. A point simply denotes any $x \in \MM$. A vertex is a point
corresponding to some vertex of the triangulation $G(\MM)$.

@@

\medskip
\noindent
{\bf Highest points $\h(c)$:} For each color $c$, we maintain a special critical point $\h(c)$ of this color.
It is supposed to denote the ``highest" point of color $c$, but the notion of highest evolves
with the algorithm.

@@

\smallskip
\begin{compactenum}
	\item If $K$ is empty, pick arbitrary $c$. Let $x$ be the highest point in $T(c)$.
	Push $x$ onto stack and set $\mcol(x) = \{c\}$.
	\item Let $h$ be head of $\st$.
	\item While $\col(h) \setminus \mcol(h)$ non-empty:
	\begin{compactenum}
		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
		\item Obtain highest point $x$ in $T(c')$. 
		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
		and set $\mcol(x) = \{c'\}$.
		\item Let $h$ be new head of $\st$.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

@@

\subsection{Constructing $\reeb(\MM)$} \label{sec:const}

The main construction procedure is $\onestep(\MM)$. It pieces together $\reeb(\MM)$
incrementally.

\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\onestep(\MM)$}

\smallskip
\begin{compactenum}
	\item Run $\update(\st)$. Pop $\st$ to get $h$.
	\item Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
	\begin{compactenum}
		\item Connect $h$ (in $\reeb(\MM)$) to $\h(C)$ and the unique minimum corresponding to split $h$. 
		\item Delete $h$ from $T(c)$, set $\h(C) = h$.
		\item End procedure.
	\end{compactenum}
	\item Let $\mcol(h) = \{c_1, c_2\}$ (so $h$ is merge).
	\item Connect $h$ in $\reeb(\MM)$ to $\h(c_1)$ and $\h(c_2)$.
	\item Delete all copies of $h$ from $T(c_1)$ and $T(c_2)$.
	\item Perform union of colors $c_1$ and $c_2$ (denote merged color as $c$). Merge heaps
	to get $T(c) = T(c_1) \cup T(c_2)$.
	\item Set $\h(c) = h$.	
\end{compactenum}
\end{minipage}}

\medskip
\begin{theorem} \label{thm:piece} Consider the very first call to $\onestep(\MM)$. Let $y$ be a point
infinitesimally below $h$ on a decreasing edge. Consider $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$ (the upper and lower manifolds). Then $\onestep(\MM)$
outputs $\reeb(\MM_u)$ and implicitly gives a valid painting of $\MM_\ell$.
\end{theorem}

We have an important lemma that helps to bound the running time.

\begin{lemma} \label{lem:tree} For every color $c$, all points in $T(c)$ lie along a monotone path in 
$\reeb(\MM)$.
\end{lemma}

\begin{proof} Prove by induction of the number of edges in $\reeb(\MM)$. Consider the head $h$ of $\st$ in the first call to $\onestep(\MM)$,
and the contour $\cont(y)$ as defined in \Thm{piece}. Take $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$. Apply induction
hypothesis on $\MM_\ell$, and use that to argue about $T(c_1)$ and $T(c_2)$ in the original manifold.
\end{proof}
